TBA

### 2020-09-23 manuscript.tex
Updates from Overleaf

%\newcommand*{\rood}[1]{\textcolor{red}{#1}}
\newcommand*{\rood}[1]{#1}
%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauw}[1]{#1}
%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauwr}[1]{#1}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Mike Jones}
\affiliation{%
  Pritzker School of Molecular Engineering, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

@@

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems.pdf, Rothemund2006FoldingPatterns, Gu2010ALine}. Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}. \noter{Please correct reference ``MhatreV.HoJi-AnnLee2012NIHAccess''} Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions. Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble meta-stable states facilitates the transition. Recent breakthrough studies have coupled experimental techniques with machine learning and MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates. The stability of certain intermediates states, such as out-of-register or  base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019}). Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}. Sandstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C bond placement in the 10-mer oligonucleotide sequence \citep{Sanstead2016}. In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to  metastable structures mentioned above.

\textcolor{purple}{I would suggest re-organizing this paragraph to emphasize how the dynamic insight from various experiments is limited and therefore high-level simulations are required to understand DNA dynamics in detail. Then, you can summarize why simulating DNA hybridization is difficult and the previously employed techniques} Given the long timescales on which DNA hybridization and dissociation events occur, the study of these processes are generally not amenable to direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling \citep{Schmitt2013ExploringSurface} transition path sampling \citep{Sambriski2009}, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Other computational studies use dramatically elevated temperature or denaturing solvent concentrations induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Experimental studies often employ temperature-jump, salt-jump, pH-jump, or other perturbative methods to drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. Single molecule diffusion and tethered multifluorophore assays facilitate equilibrium analysis, however these present technical difficulties and are hampered by data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices}. Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}.

In this work, we use the coarse-grain 3 site per nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.  We leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that GC base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of meta-stable states during the hybridization and dissociation process. Furthermore we can perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (shorter lag time) and generate more detailed models. Additionally, we can compare slow dynamical modes and meta-stable states between sequence-specific SRV-MSMs. Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequences dependent dynamics, meta-stable structures, and relative timescales.

@@

Is some additional information warranted on the model itself before going into this specific setup?

@@

We initialized four sequences previously investigated by Sandstead et al. -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}. We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 \textcolor{purple}{So these are initially in the bulk and then may interact with the DNA? Are these the effective concentrations after you shrink the box?} were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}. In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A). This translated to a box size of 77.74 Å and an effective oligo concentration of 7 mM. We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions. We maintained a Debye-Huckel screening potential with an ionic strength of 240 mM -- corresponding to NaCl concentration -- to account for phosphate backbone interactions. These electrostatics preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}. We run our simulations in the NVE ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. Simulation temperature was determined by sequence specific melting temperature such that strands were equally likely to spend time in the hybridized or dissociated state. Tmelt simulations were run for $3*10^{8}$ time steps, equating to 4.5$\mu$s simulation time, and frames were saved every 30 ps. A 15 fs timestep was used for all simulations runs. For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation. An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state. In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed  \textcolor{purple}{Why 1.5 microseconds? Is this arbitrary or is there a good reason to remove 1.5?}, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Based on the self-complementary nature of each sequence, we averaged permutable distances (45 pairs in total) together. This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method \textcolor{purple}{Since the VAMP-2 scoring comes up a lot, can you maybe another sentence or two on how exactly this works? Citing the Noe paper helps, but I think experimentalists may get a bit lost with it}, which represents the sum of squared estimates for the transfer operator, was employed to evaluate the quality of this feature set \citep{Mardt2018VAMPnetsKinetics}. We compared the VAMP-2 score for the permutation-free distances to the complete set of intermolecular distances and found only a small loss in kinetic variance when using the reduced data set. Furthermore, we found an increase in VAMP-2 score when using reciprocal pairwise distances and chose these reciprocal permutation-free coordinates as a consistent feature set. These features were normalized and passed into sequence-specific SRVs.

@@

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features. This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator can be estimated (equation here?). The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs. Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Harrigan2017LandmarkDynamics}. SRVs provide robust nonlinear approximation and computation timescales linearly with the amount of input data. This is a key attribute to our system as 10 million frames with 55 features in each frame are used as input data. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen}, \citep{Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

Using optimized hyperparameters and featurized trajectory data, we generated a hierarchical dynamic encoder (HDE) to transform 55 reciprocal pairwise distances into a low dimensional SRV feature set. In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes. We determined the number of slow mode via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}. In particular we looked for a diminishinghttps://www.overleaf.com/project/5eff37185f425e0001a920ca increase in the vamp-2 score, inconsistency between cross validation scores -- suggesting that the model maybe learning different modes, and a widening gap between train and test scores -- suggesting that the model might be overfitting on statistical noise in the training data. We used a batch size of 50000 and ran each model for total of 30 training epochs. We used two hidden layers and set the size of each layer to 100. For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training. SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs. SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com} \citep{Abadi2016TensorFlow:Systems}.

@@

MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way. The technique relies on the discretitization of kinetically similar conformations into microstates and finds the conditional probability between states within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. This becomes crucial in our system given the large difference in timescales between leading modes. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis. To prove this, SRV-MSM VAMP-2 scores were shown to be consistently higher than MSMs constructed from TICA coorinates (TICA-MSMs) \citep{Sidky}. Furthermore, SRV-MSM implied timescales converge faster than TICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model. To build our SRV-MSM framework, we employed the Pyemma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. After passing in SRV coordinates, we perform k-means microstate clustering, Bayesian MSM construction, and PCCA+ macrostate assignments. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results.

@@

Beyond state probabilities and free energy approximations, the macrostate MSM also yields discrete kinetic information in the form of transition probabilities between states. Figure \ref{fig:AT-all_in_one} shows the probability of moving from one state to another within the MSM lag time (1.2 ns). All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation \textcolor{purple}{This is consisting with the GCGCGC sequence of Wales, right? They saw a large barrier to association from the double shifted state}. In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state. Although there is an approximately equal probability of transitioning from the dissociated state to the hybridized and 5' shifted states, there is a 2x lower probability of moving to the 3' shifted states. These asymmetric kinetic effects may have some role in the substantially lower state populations we see above. From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state \textcolor{purple}{Based on the thermo studies of dangling ends and inert tails, it seems like the role of shifted states may depend on salt conditions, this could be an interesting future direction.}.

>>>>

%\newcommand*{\rood}[1]{\textcolor{red}{#1}}
\newcommand*{\rood}[1]{#1}
%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauw}[1]{#1}
%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauwr}[1]{#1}
\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

@@

\usepackage[normalem]{ulem}
\usepackage{algorithm}
\usepackage{subfig}
\usepackage{algpseudocode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Michael S. Jones}
\affiliation{%
  Pritzker School of Molecular Engineering, %
  The University of Chicago, %
  929 East 57th Street, Chicago, Illinois 60637, United States%
}

@@

\begin{comment}
%%%%%% intro notes %%%%%%%
\citep{Sanstead2018DirectDehybridization} Good intro with refs and focus on importance of dynamics "rich dynamics even for short oligos" ... Many of the most important functions performed by DNA are highly dynamic, whether in the the biological roles of replication transcription, and gene regulation, or in the expanding field of DNA-based nanotechnology
\end{comment}

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems.pdf, Rothemund2006FoldingPatterns, Gu2010ALine}. Recently, structural DNA nanotechnology has enabled self-assembly on micro to milli scales, and dynamic DNA nanotechnology has been used to perform basic calculation and to probe single molecules via temporal DNA signatures\citep{MhatreV.HoJi-AnnLee2012NIHAccess, Bui2018, Shah2019}. Both technologies rely on the hybridization reaction between complementary DNA strands and leverage the flexibility of shorter DNA oligomers to participate in these reactions. Although many experimental and computational studies have rigorously explored DNA dynamical phenomena such as hybridization, hairpin formation, and single base pair flipping, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble meta-stable states facilitates the transition. Recent breakthrough studies have coupled experimental techniques with machine learning and MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we broaden our analysis into higher order dynamical processes and meta-stable intermediates. The stability of certain intermediates states, such as out-of-register or shifted base pairing in repetitive sequences, has been well documented in previous computational studies \citep{Phys2014, Xiao2019, Maciejczyk2014DNAModel}). Furthermore, frayed structures and dynamics have been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}. Sandstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}. In this work, we study the same four sequences explored by Sandstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above.

Our understanding of DNA dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these present technical difficulties and are hampered by data collection rates \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices}. Förster resonance energy transfer (FRET) analysis, particularly when coupled with methods mentioned above, provides additional resolution, but it is unclear how fluorescent tags may interfere with the dynamics of short nucleotides \citep{Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several high-level computational models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all atom simulations, the long timescales on which DNA hybridization and dissociation events occur make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}.Other computational works use dramatically elevated temperature or denaturing solvent concentrations induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied certain aspect the overall dynamics process in one direction.

\textcolor{purple}{Could add more here about all-atom simulations?}

In this work, we use the coarse-grain 3 Sites Per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}.  We leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstates clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that GC base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of meta-stable states during both the hybridization and dissociation process. Furthermore we can perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (shorter lag time) and generate more detailed models. Additionally, we can compare slow dynamical modes and meta-stable states between sequence-specific SRV-MSMs. Within these meta-stable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant DNA dynamics \textcolor{purple}{Not sure how much to say about this here, but I think these experiments should be mentioned in the intro}. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequences dependent dynamics, meta-stable structures, and relative timescales.

@@

\textcolor{purple}{Is some additional information warranted on the model itself before going into this specific setup?}

@@

We initialized four sequences previously investigated by Sandstead et al. -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands according to 3SPN.2 documentation \citep{Sanstead2016, Phys2014}. We initialized explicit ions such that 240 mM NaCl and 18 mM MgCl2 were added to the box in addition to 18 Na counter ions to balance the charge from the 9 phosphate groups in each oligonucleotide backbone \citep{Hinckley2015}. In order to maximize concentration without allowing strands to see each other through periodic boundaries, we set the box size just larger than the sum of the maximum end-to-end extension length of a single strand and the force cutoff (using Ewald summation method is set at 20 A). This translated to a box size of 77.74 Å and an effective oligo concentration of 7 mM. We used an Ewald potential to calculate long range Coulombic interaction between DNA and ions. We used a Debye-Huckel screening potential to account for phosphate-backbone interactions and preserve the persistence length and intrinsic curvature of DNA while taking the effects of ion-DNA interactions\citep{Hinckley2015}. We ran our simulations in the NVT ensemble and fix temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. We set the simulation temperature was set about 1 K higher than empirical sequence-specific melting temperature in order to maximize transitions between dissociated and hybridized states. We use a 15 fs time step and $3*10^{8}$ steps in each simulation, equating to 4.5$\mu$s simulation time. Base pair coordinates were saved every 30 ps, and full frames including backbones and ions were saved every 3 ns for visualization purposes. For each sequences, 100 simulations were performed in parallel, consuming about 32 serial cpu-hours of computation time for each independent simulation. An approximate Tmelt Boltzman distribution was replicated by initializing half of runs from the hybridized state and half from a random dissociated state. In order to allow for further equilibration, the first third (1.5 $\mu$s) of each simulation was removed, resulting in 100 x 100000 frames and a total of 300$\mu$s simulation time per sequence.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Based on the self-complementary nature of each sequence, we averaged permutable distances (45 pairs in total) together. This permutation reduction follows a similar procedure used in TICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system (equation below), providing a robust and object means to evaluate various parameters and models. In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$. The score $R_2$ can then be found for $\chi$ by applying the VAMP principle with cross-validation.

\begin{align}
 	\mathscr{C}_{00}=&\Expect{\chi(x_t)\chi(x_t)^\intercal}_t\\
 	\mathscr{C}_{01}=&\Expect{\chi(x_t)\chi(x_{t+\tau})^\intercal}_t\\
 	\mathscr{C}_{11}=&\Expect{\chi(x_{t+\tau})\chi(x_{t+\tau})^\intercal}_{t+\tau}\\
	\\
 	R_2[\chi]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2 +1
 	%R_2^{val}[\chi]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1
\end{align}\label{CK1}

% for more general case with  \chi_0, \chi_1, but in practice this is not used
\begin{comment} 
\begin{align}
 	\mathscr{C}_{00}=&\Expect{\chi_0(x_t)\chi_0(x_t)^\intercal}_t\\
 	\mathscr{C}_{01}=&\Expect{\chi_0(x_t)\chi_1(x_{t+\tau})^\intercal}_t\\
 	\mathscr{C}_{11}=&\Expect{\chi_1(x_{t+\tau})\chi_1(x_{t+\tau})^\intercal}_{t+\tau}\\
	\\
 	R_2[\chi_0, \chi_1]=&\norm{\mathscr{C}_{00}^{-1/2}\mathscr{C}_{01}\mathscr{C}_{11}^{-1/2}}_F^2\\
 	R_2^{val}[\chi_0, \chi_1]=&\norm{\mathscr{(C_{00}^{val})}^{-1/2}\mathscr{C}_{01}^{val}\mathscr({C}_{11}^{val})^{-1/2}}_F^2 + 1\\
\end{align}\label{CK2}
\end{comment}

We compared scores for the permutation-free distances, intermolecular, and intra+intermolecular distances and found small differences between the three that varied by sequences (show all sequences in supplemental), but overall we did not observe a loss in kinetic information when using the reduced feature set. We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal permutation-free coordinates as a consistent feature set. These features were normalized and passed into sequence-specific SRVs.

% include covariance, vamp-2  and validation equations, lag time equation from vampnet paper mardt2018

@@

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The framework uses a twin-lobed artificial neural network, similar in structure to the network employed in VAMPnets \citep{Mardt2018VAMPnetsKinetics}, to learn a low dimensional representation of input features. This representation is optimized for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs. Other methods, such as time-lagged independent component analysis (TICA) and kernel TICA (kTICA), have been employed in place of SRVs but suffer limitations from the linear transformation in the case of the former and high computational cost in the case of the latter \citep{Molgedey1994SeparationCorrelations, Harrigan2017LandmarkDynamics}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data. This is a key attribute to our system as 10 million frames with 55 features in each frame are used as input data. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

Using optimized hyperparameters and featurized trajectory data, we transform 55 reciprocal pairwise distances into a low dimensional SRV basis set. In order to maintain consistency between sequences, we kept all SRV training hyperparameters the same with the exception of the number of outputed slow modes. We determined the number of slow modes via cross-validation on the VAMP-2 score to ensure that the coordinate did not over fit on statistical noise \citep{McGibbon2015VariationalKinetics}. In particular, we looked for convergence in the vamp-2 score and inconsistency between cross validation scores -- suggesting that the model may be overfitting on artifacts in the training data. We used a batch size of 50000 and ran each model for a total of 30 training epochs. We used two hidden layers and set the size of each layer to 100. For cross-validation and comparison between different hyper-parameters, we used a 80/20 validation split training. SRV training required about 22 GPU-minutes across 1 GPU and 10 CPUs. SRV training was implemented using Keras and Tensorflow \citep{KerasGithub.Com, Abadi2016TensorFlow:Systems}.

@@

MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way. The technique relies on the discretitization of kinetically similar conformations into microstates and finds the conditional probability between states within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. This becomes crucial in our system given the large difference in timescales between leading modes. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis. To prove this, SRV-MSM VAMP-2 scores were shown to be consistently higher than MSMs constructed from TICA coorinates (TICA-MSMs) \citep{Sidky}. Furthermore, SRV-MSM implied timescales converge faster than TICA-MSM timescales, enabling a shorter lag time and therefore a higher resolution model. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. After passing in SRV coordinates, we perform k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results.

@@

Beyond state probabilities and free energy approximations, the macrostate MSM yields valuable discrete kinetic information in the form of transition probabilities between states. Figure \ref{fig:AT-all_in_one} shows the probability of moving from one state to another within the MSM lag time (1.2 ns). All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state. Although there is an approximately equal probability of transitioning from the dissociated state to the hybridized and 5' shifted states, there is a 2x lower probability of moving to the 3' shifted states. These asymmetric kinetic effects may have some role in the substantially lower state populations we see above. From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state. We recognize that these effects could be sensitive to the specific ion environment -- especially considering the effect of ionic strength on inert tails -- and warrant further computational and experimental investigation beyond this work.
